---
unit: "ETC3450"
unittitle: "Time Series Econometrics"
title:  "Assignment 2"
subtitle: "Housing Data: Time Series Properties"
author:
- name: Ari Gestetner
  email: ages0001@student.monash.edu
- name: Nathan Giofkou
  email: ngio0003@student.monash.edu
- name: Mitchell Evans 
  email: meva0009@student.monash.edu
bibliography: references.bib
format:
  html:
    execute:
      echo: true
    code-fold: true
    code-line-numbers: true
    code-copy: hover
  report-pdf:
    execute:
      echo: false
toc: true
execute: 
  warning: false
---

```{r}
#| eval: false

renv::load("Assignment2")
setwd("Assignment2")
```  

```{r}
library(tidyverse)
library(tsibble)
library(feasts)
library(urca)
library(tseries)
library(egcm)
library(vars)
library(dplyr)

theme_set(theme_bw(base_size = 16))

housingdata <- read_csv("./housingdata.csv") |>
  tibble() |>
  mutate(date=yearquarter(date)) |>
  as_tsibble()
```  


# Stationarity: 

> Determine whether series are stationary (use visualisation and hypothesis tests)

```{r}
#| fig-pos: 'H'
#| title: Plot data
#| layout-ncol: 2
#| layout-nrow: 2

housingdata |> 
  ggplot(aes(x=date,y=house_mel)) +
    labs(x="Date", y="Price", title='Melbourne Median House Prices Over Time') +
    geom_line()

housingdata |> 
  ggplot(aes(x=date,y=appt_mel)) + 
    labs(x="Date", y="Price", title='Melbourne Median Apartment Prices Over Time') + 
    geom_line()

housingdata |> 
  ggplot(aes(x=date,y=house_syd)) + 
    labs(x="Date", y="Price", title='Sydney Median House Prices Over Time') + 
    geom_line()

housingdata |> 
  ggplot(aes(x=date,y=appt_syd)) + 
    labs(x="Date", y="Price", title='Sydney Median Apartment Prices Over Time') + 
    geom_line()
```

Visually, all series look non-stationary as their means are clearly time dependent and all series exhibit an upwards trend. Doesn’t seem to be any seasonality.

## Formal Hypothesis tests

### Melbourne Housing Prices

```{r}
#| fig-pos: 'H'
#| label: fig-melb-acf-pacf 
#| fig-cap: "ACF and PACF plots of Melbourne housing prices"
#| fig-subcap:
#|   - "The ACF has very slow decay, indicating non stationarity."
#|   - "Significant spike at lag $1$ in PACF indicates we use $1$ lag for an ADF test."
#| layout-ncol: 2

ACF(housingdata, house_mel) |>
  autoplot() +
  labs(x="Lag", y="ACF", title="ACF of Melbourne Housing data")
PACF(housingdata, house_mel) |>
  autoplot() +
  labs(x="Lag", y="PACF", title="PACF of Melbourne Housing data")

# ADF test (hm_ stands for homes melbourne)
hm_adf_result <- ur.df(housingdata$house_mel, type = "drift", selectlags = "AIC") 
hm_adf_result_sum <- summary(hm_adf_result)

hm_adf_teststat <- round(hm_adf_result_sum@teststat[1], 2)
hm_adf_cval <- round(hm_adf_result_sum@cval[1,2], 2)

#KPSS test
hm_kpss_result <- kpss.test(na.omit(housingdata$house_mel))
hm_kpss_pval <- round(hm_kpss_result$p.value, 2)
```


For the ADF test: The $\tau$ statistic was given as $`r hm_adf_teststat`$, which is greater than the critical value of $`r hm_adf_cval`$ at the $5\%$ significance level. Thus, we fail to reject $H_0$ and conclude that the series may have a unit root and differencing is necessary.

For the KPSS test: the $p$-value was given as $`r hm_kpss_pval`$, so we reject the null in favour of the series being non stationary, which supports the conclusion from the ADF test.

From the PACF, the strong spike at lag $1$ suggests the presence of persistence in the series, which is consistent with non-stationary behaviour. Combined with formal unit root tests, this indicates that the series likely becomes stationary after first order differencing. Therefore, the Melbourne housing series is likely $I(1)$.


### Melbourne Apartment Prices

```{r}
#| fig-pos: 'H'
#| layout-ncol: 2

ACF(housingdata, appt_mel) |>
  autoplot() +
  labs(x="Lag", y="ACF", title="ACF of Melbourne Apartment data")
PACF(housingdata, appt_mel) |>
  autoplot() +
  labs(x="Lag", y="PACF", title="PACF of Melbourne Apartment data")

# ADF test (am_ stands for apartments melbourne)
am_adf_result <- ur.df(housingdata$appt_mel, type = "drift", selectlags = "AIC") 
am_adf_result_sum <- summary(am_adf_result)

am_adf_teststat <- round(am_adf_result_sum@teststat[1], 2)
am_adf_cval <- round(am_adf_result_sum@cval[1,2], 2)

#KPSS test
am_kpss_result <- kpss.test(na.omit(housingdata$appt_mel))
am_kpss_pval <- round(am_kpss_result$p.value, 2)
```  

For the ADF test: The $\tau$ statistic was given as $`r am_adf_teststat`$, which is greater than the critical value of $`r am_adf_cval`$ at the $5\%$ significance level. Thus, we fail to reject $H_0$ and conclude that the series may have a unit root and differencing is necessary.

For the KPSS test: the $p$-value was given as $`r hm_kpss_pval`$, so we reject the null in favour of the series being non stationary, which supports the conclusion from the ADF test.

From the PACF, the strong spike at lag $1$ suggests the presence of persistence in the series, which is consistent with non-stationary behaviour. Combined with formal unit root tests, this indicates that the series likely becomes stationary after first differencing. Therefore, the Melbourne apartment series is likely $I(1)$.


### Sydney Housing Prices  

```{r}
#| fig-pos: 'H'
#| layout-ncol: 2

ACF(housingdata, house_syd) |>
  autoplot() +
  labs(x="Lag", y="ACF", title="ACF of Sydney Housing data")
PACF(housingdata, house_syd) |>
  autoplot() +
  labs(x="Lag", y="PACF", title="PACF of Sydney Housing data")

# ADF test (hs_ stands for houses sydney)
hs_adf_result <- ur.df(housingdata$house_syd, type = "drift", selectlags = "AIC") 
hs_adf_result_sum <- summary(hs_adf_result)

hs_adf_teststat <- round(hs_adf_result_sum@teststat[1], 2)
hs_adf_cval <- round(hs_adf_result_sum@cval[1,2], 2)

#KPSS test
hs_kpss_result <- kpss.test(na.omit(housingdata$house_syd))
hs_kpss_pval <- round(hs_kpss_result$p.value, 2)
```


For the ADF test: The $\tau$ statistic was given as $`r hs_adf_teststat`$, which is greater than the critical value of $`r hs_adf_cval`$ at the $5\%$ significance level. Thus, we fail to reject $H_0$ and conclude that the series may have a unit root and differencing is necessary.

For the KPSS test: the $p$-value was given as $`r hs_kpss_pval`$, so we reject the null in favour of the series being non stationary, which supports the conclusion from the ADF test.

From the PACF, the strong spike at lag $1$ suggests the presence of persistence in the series, which is consistent with non-stationary behaviour. Combined with formal unit root tests, this indicates that the series likely becomes stationary after first differencing. Therefore, the Sydney housing series is likely $I(1)$.



### Sydney Apartment Prices

```{r}
#| fig-pos: 'H'
#| layout-ncol: 2

ACF(housingdata, appt_syd) |>
  autoplot() +
  labs(x="Lag", y="ACF", title="ACF of Sydney Apartment data")
PACF(housingdata, appt_syd) |>
  autoplot() +
  labs(x="Lag", y="PACF", title="PACF of Sydney Apartment data")

# ADF test (as_ stands for apartments sydney)
as_adf_result <- ur.df(housingdata$appt_syd, type = "drift", selectlags = "AIC") 
as_adf_result_sum <- summary(as_adf_result)

as_adf_teststat <- round(as_adf_result_sum@teststat[1], 2)
as_adf_cval <- round(as_adf_result_sum@cval[1,2], 2)

#KPSS test
as_kpss_result <- kpss.test(na.omit(housingdata$appt_syd))
as_kpss_pval <- round(as_kpss_result$p.value, 2)
```


For the ADF test: The $\tau$ statistic was given as $`r as_adf_teststat`$, which is greater than the critical value of $`r as_adf_cval`$ at the $5\%$ significance level. Thus, we fail to reject $H_0$ and conclude that the series may have a unit root and differencing is necessary.

For the KPSS test: the $p$-value was given as $`r as_kpss_pval`$, so we reject the null in favour of the series being non stationary, which supports the conclusion from the ADF test.

From the PACF, the strong spike at lag $1$ suggests the presence of persistence in the series, which is consistent with non-stationary behaviour. Combined with formal unit root tests, this indicates that the series likely becomes stationary after first differencing. Therefore, the Sydney apartment series is likely $I(1)$.


## Conclusion on Stationarity:
All series are determined to be non-stationary and have unit roots. All are integrated of order $1$-$I(1)$; meaning that taking the first difference of each series should make each series stationary.

# Cointegration

> Pairs:
  - Houseprices in Melbourne and Sydney  
  - Apartment prices in Melbourne and Sydney  
  - Apartment prices in Melbourne and house prices in Melbourne  
  - House prices in Sydney and house prices in Sydney  

## Test for cointegration of each pair  
Use Engle-Granger procedure as dealing with pairs  
1. Run long run regression $y_t = a + \beta x_t + u_t$  
2. Save $u_t$  
3. Test the residuals for stationarity (ADF test on ut)  
	$H_0$: resids have a unit root (no cointegration)  
	$H_1$: resids are stationary (cointegration)  

### Houseprices in Melbourne and Sydney  

```{r}
hm_hs_test <- housingdata |> 
  lm(house_mel ~ house_syd, data = _) |>
  residuals() |>
  adf.test()

hm_hs_p_val <- hm_hs_test |> pluck("p.value") |> round(2)
```

$p$-value was given as $`r hm_hs_p_val`$ meaning there is insufficient evidence to reject $H_0$, indicating that the residuals may be non stationary. Thus, there is no evidence of cointegration between house prices in melbourne and sydney 

### Apartment prices in Melbourne and Sydney  
```{r}
am_as_test <- housingdata |> 
  lm(appt_mel ~ appt_syd, data = _) |>
  residuals() |>
  adf.test()

am_as_p_val <- am_as_test |> pluck("p.value") |> round(2)
```  

$p$-value was given as $`r am_as_p_val`$ meaning there is insufficient evidence to reject $H_0$, indicating that the residuals may be non stationary. Thus, there is no evidence of cointegration between apartment prices in melbourne and sydney 

### Apartment prices in Melbourne and house prices in Melbourne  
```{r}
am_hm_test <- housingdata |> 
  lm(house_mel ~ appt_mel, data = _) |>
  residuals() |>
  adf.test()

am_hm_p_val <- am_hm_test |> pluck("p.value") |> round(2)
```  

$p$-value was given as $`r am_hm_p_val`$ meaning there is insufficient evidence to reject $H_0$, indicating that the residuals may be non stationary. Thus, there is no evidence of cointegration between apartment prices and house prices in melbourne 

### Apartment prices in Sydney and house prices in Sydney  
```{r}
as_hs_test <- housingdata |> 
  lm(house_syd ~ appt_syd, data = _) |>
  residuals() |>
  adf.test()

as_hs_p_val <- as_hs_test |> pluck("p.value") |> round(2)
```  

$p$-value was given as $`r as_hs_p_val`$ meaning there is insufficient evidence to reject $H_0$, indicating that the residuals may be non stationary. Thus, there is no evidence of cointegration between apartment prices and house prices in Sydney

### Conclusion for pair wise cointegrations  
No cointegration between the given pairs, meaning the two price series only move together in the short run and don’t move around a common long run equilibrium.  

## Does one (or more) cointegrating relationship exist among all four variables  
### Johansen Test  
Following the PACF’s, we will want one lag in differences so for the Johansen test we use K = 2

```{r}
#| eval: false
housing_jo_summ <- housingdata |> 
  tibble() |> 
  select(house_mel, appt_mel, house_syd, appt_syd) |> 
  ca.jo(
    ecdet = "const", 
    type="eigen", 
    K = 2, 
    spec="longrun", 
    season=4) |> 
  summary()

# Transform Johansen test results into readable table
housing_jo_summ |> pluck("cval") |>
  cbind(
    housing_jo_summ |> pluck("teststat")
  ) |>
  as.data.frame() |>
  setNames(c("10pct", "5pct", "1pct", "test")) |>
  mutate(
    hypothesis = c("r <= 3", "r <= 2", "r <= 1", "r = 0"),
    .before = test
  ) |>
  select(hypothesis, test, `10pct`, `5pct`, `1pct`) |>
  tibble() |>
  knitr::kable(
    col.names = c("Test", "Statistic", "10%", "5%", "1%"),
    digits = 2,
    caption = "Johansen Cointegration Test Results"
  )
```

As the test is not rejected at any level, we can say that there are no cointegrating relationships 


## Conclusion From Johansen Test  
No cointegrating relationships were found, which matches the results from the Engle Granger tests in which the tested pairs did not have a cointegrating relationship, thus while there may be short run relationships, there are no stable long-run equilibrium relationships. An explanation could be due to city-specific or property-type factors.

# Models for the Data  

> Model the data using one or more of 
  - VAR model  
  - Error correction model  
  - ARDL model  
  - VECM model  

## VAR Model  

We will be using a VAR model on the first differences of the series. This is due to all series being $I(1)$ and there being no evidence of cointegration, meaning that there are no long run relationships that have to be modelled with a VECM.

This lets us: 
- Capture short run interdependence
- Evaluate how shocks to one market affects others


$$
y_{t}=\begin{bmatrix}
\Delta \texttt{house\_mel} \\
\Delta \texttt{house\_syd} \\
\Delta \texttt{appt\_mel} \\
\Delta \texttt{appt\_syd}
\end{bmatrix}
$$  

And the model is  
$$
\mathbf{y_{t}} = \mathbf{c}+\mathbf{\Phi y_{t-1}}+\mathbf{\varepsilon_{t}}
$$

Where $\mathbf{c}$ is a $4\times 1$ intercept vector and $\mathbf{\Phi}$ is a $4\times 4$ coefficient matrix.  

```{r}
var_housingdata <- housingdata |>
  mutate(
    d_house_mel = c(NA, diff(house_mel)),
    d_appt_mel = c(NA, diff(appt_mel)),
    d_house_syd = c(NA, diff(house_syd)),
    d_appt_syd = c(NA, diff(appt_syd))
  ) |>
  drop_na()

var_housing <- var_housingdata |>
  tibble() |>
  dplyr::select(d_house_mel, d_appt_mel, d_house_syd, d_appt_syd) |>
  VAR(p=1, type = "const")

var_housing |> irf(
    # The variable receiving the shock
    impulse = "d_appt_mel",  
    # Affected variables
    response = c("d_house_syd", "d_appt_syd", "d_house_mel", "d_appt_mel"),
    # Number of periods to trace the effect (e.g. 12 quarters = 3 years)
    n.ahead = 4,  
    ) |>
    plot()
```

